A Detailed Demographic Analysis of 2020 U.S. Presidential Election Through Multilevel Regression and Post-Stratification

### 2024-03-16 paper/paper.qmd
final commit!

# education plot
  ggplot(vote_trump_education, aes(y = reorder(education, trump_predict), 
                                   x = trump_predict)) +
  geom_point(stat = "identity") +
  geom_errorbar(aes(xmin = trump_prop_lower, xmax = trump_prop_upper, width = 0.2)) +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "green") +
  scale_x_continuous(labels = scales::percent) +
  labs(title = "Estimated Proportion of Voters for Trump by Education",
       x = "Percentage", y= "Education") + 
  theme_minimal() + theme(axis.text.y = element_text(size = 12))

@@

The analysis reveals critical insights into how education, race, gender, and age influence voting behavior. Contrary to conventional wisdom and past electoral analyses, our findings indicate a divergent pattern in voting behavior among different educational groups. The shift in party allegiance among non-college voters from Republicans to Democrats and the dominance of popularity of the Republicans with 'Some college' and 'Postgrad' education levels challenge the stereotypical narratives that educational attainment leads to an increase in support for the Democrats. This group, especially the highly educated 'Postgrad' group being the second highest supporters suggests a reevaluation of Democrats' strategies to appeal to educated voters concerned with economic policies and national security. However, according to our analysis, the Republicans failed to attract voters as the highest proportion of support among the four educational groups remains under 40%. This almost certainly predicts that Joe Biden will become the president of the United States. This coincides with the election results where Biden won 55% of the votes from the college graduates while Trump gained 43% [@theWashingtonPost]. Our analysis of 37% of voters with college degrees and 37.5% of postgrad suggests this is in line with the results. However, this model has failed to estimate the winning of Trump from the less educated group since we estimated 39% from voters with 'some college degree' and 36.5% from voters with 'High school or less' whereas the post-election results show 50% for Trump from 'Some college or less' [@theWashingtonPost].

@@

We were able to estimate the winning states for each candidate with some success. From the data and analysis outlined above, we believe that our model suggests that Biden would have won both the popular vote and the electoral votes and become the president of the United States. This was partially right in that he did win both but with a smaller margin than we had expected. In the end, he won 306 electoral votes and won 51.3% of the votes to become the 46th US president [@CNN]. however, + add why this happened

@@

Future research should explore models that can accommodate multiple electoral outcomes, including third-party voting and more variables if possible. Using more accurate or applicable real-time data sources could complement traditional survey methods and offer fresh insights into voter preferences. Societal norms and political landscapes change over time. Therefore, future research should focus on emerging social issues that may influence voter behavior. In the 2024 Presidential election, climate change, technology policy, and social justice could play more significant roles in shaping electoral outcomes.

>>>>

# education plot
  ggplot(vote_trump_education, aes(y = reorder(education, trump_predict), 
                                 x = trump_predict)) +
  geom_point(stat = "identity") +
  geom_errorbar(aes(xmin = trump_prop_lower, xmax = trump_prop_upper, width = 0.2)) +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "green") +
  scale_x_continuous(labels = scales::percent) +
  labs(title = "Estimated Proportion of Voters for Trump by Education",
       x = "Percentage", y= "Education") + 
  theme_minimal() +
  theme(axis.text.y = element_text(size = 12),
        plot.title = element_text(hjust = 2)) # Move title to the right
```

@@

The analysis reveals critical insights into how education influence voting behavior. Contrary to conventional wisdom and past electoral analyses, our findings indicate a divergent pattern in voting behavior among different educational groups. The shift in party allegiance among non-college voters from Republicans to Democrats and the dominance of popularity of the Republicans with 'Some college' and 'Postgrad' education levels challenge the stereotypical narratives that educational attainment leads to an increase in support for the Democrats. This group, especially the highly educated 'Postgrad' group being the second highest supporters suggests a reevaluation of Democrats' strategies to appeal to educated voters concerned with economic policies and national security. However, according to our analysis, the Republicans failed to attract voters as the highest proportion of support among the four educational groups remains under 40%. This almost certainly predicts that Joe Biden will become the president of the United States. This coincides with the election results where Biden won 55% of the votes from the college graduates while Trump gained 43% [@theWashingtonPost]. Our analysis of 37% of voters with college degrees and 37.5% of postgrad suggests this is in line with the results. However, this model has failed to estimate the winning of Trump from the less educated group since we estimated 39% from voters with 'some college degree' and 36.5% from voters with 'High school or less' whereas the post-election results show 50% for Trump from 'Some college or less' [@theWashingtonPost]. This is odd because the higher the education completed, the more likely an individual to vote for democratic party in general. One possible explanation for this is that there are some issues with the data collection where voting for Trump is underestimated as the the proportion of voters for him remain under 40% across all educational groups. Another possibility is that the gap between educational groups in party preference has actually been reduced from the past. In fact, support from white men without a college degree reduced somewhat although he still won a majority of this demographic [@pewresearchcenter1].

@@

We were able to estimate the winning states for each candidate with some success. From the data and analysis outlined above, we believe that our model suggests that Biden would have won both the popular vote and the electoral votes and become the president of the United States. This was partially right in that he did win both but with a smaller margin than we had expected. In the end, he won 306 electoral votes and won 51.3% of the votes to become the 46th US president [@CNN]. However, our model underestimates the proportion of Trump voters around 8 - 10%, compared to the actual 48% of votes he got in the election. In the cleaning process, we had to reduce both the number of survey data and post-stratification data, that could have led to this difference.

@@

Our analysis also involves some weaknesses of using predictive modeling to fully grasp the complexity of how elections work. Although our model picks up on some trends and changes in voter demographics, it also shows that predicting election outcomes accurately, particularly in states where the results could go either way, is difficult. The differences between what our model predicted and the actual election results underline the importance of continuously improving our models. This means adding more detailed data and possibly trying different methods to better understand how people vote. This experience has made it clear that to make our election predictions more accurate and trustworthy in the future, we need to take into account various factors, such as education levels, differences between urban and rural areas, and the evolving political scene. Therefore, future research should explore models that can accommodate multiple electoral outcomes, including third-party voting and more variables if possible. Using more accurate or applicable real-time data sources could complement traditional survey methods and offer fresh insights into voter preferences. Societal norms and political landscapes change over time. Therefore, future research should focus on emerging social issues that may influence voter behavior. In the 2024 Presidential election, climate change, technology policy, and social justice could play more significant roles in shaping electoral outcomes.
